---
title: "Lab 7: Support Vector Machines"
author: "Juliana Vinluan"
date: "2025-04-22"
output:
  html_document:
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction Questions

Briefly describe the problem the authors are trying to solve.
	•	How many predictor variables were used?
	•	What was the ratio of training data to test data?
	•	What kernels did the authors evaluate and how did they tune them?
	•	Although overall accuracy was similar in several of the kernels, the linear 
  kernel was strongly preferred. Why?

- The authors are addressing the problem of predicting the purchase of sessions 
in a web store, by exploring the use of SVMs specifically to predict 
customer purchasing behavior online.
- 23 Variables were used
- 26,000 : 13,000
- The authors evaluated three kernels: linear, polynomial, radial and sigmoid.
- The error rate was the lowest by far for the linear kernel and it was most effective as stated by the authors.

### Load the Vehicle dataset

```{r load-packages}
# Load
library(mlbench)
library(e1071)
library(caret) # For splitting
library(dplyr) # Data manipulation

# Load the Vehicle data
data(Vehicle)

# Look at the data
str(Vehicle)
summary(Vehicle)
sum(is.na(Vehicle)) #tells us about missing values

```

## Step 1 : Preprocessing the Data
1. Load the Vehicle dataset from the mlbench package.
2. Check for missing values.
3. Check the structure of the data.
4. Applied standardization (scale()) to predictor variables.
5. Put the standardized predictors together with the original Class label into vehicle_processed.
```{r scale-factors}
# Save the target variable
vehicle_target <- Vehicle$Class

# Scale predictors
vehicle_scaled <- scale(Vehicle %>% select(-Class))

# Recombine predictors + target
vehicle_processed <- data.frame(vehicle_scaled, Class = vehicle_target)

# Check result
head(vehicle_processed)

```

## Step 2: Splitting the Data
1. Used createDataPartition() from the caret package to create five different random training/test splits.
2. Each split uses approximately 70% of the data for training and 30% for testing.
```{r }
set.seed(123)  # Set a seed for reproducibility

# Create empty lists to hold training and testing sets
train_list <- list()
test_list <- list()

# Create 5 different splits
for (i in 1:5) {
  # Make indices for 70% training
  train_indices <- createDataPartition(vehicle_processed$Class, p = 0.7, list = FALSE)
  
  # Split the data
  train_data <- vehicle_processed[train_indices, ]
  test_data <- vehicle_processed[-train_indices, ]
  
  # Save them
  train_list[[i]] <- train_data
  test_list[[i]] <- test_data
}

```

```{r check}
# Check sizes
sapply(train_list, nrow)
sapply(test_list, nrow)

# Should be about 70% and 30% of the full dataset (846 rows total)
nrow(vehicle_processed)  # Should be 846
```

## Step 3: Training the Models
1. Split using 70% of the data for training and 30% for testing.
2. Save splits into train_list and test_list to be used later.
```{r train-models}
# Create empty lists to hold models
# Create empty lists to hold models separately by kernel
svm_models_linear <- list()
svm_models_poly <- list()
svm_models_rbf <- list()

# Train models for each split
for (i in 1:5) {
  train_data <- train_list[[i]]
  
  svm_models_linear[[i]] <- svm(Class ~ ., data = train_data, kernel = "linear")
  svm_models_poly[[i]] <- svm(Class ~ ., data = train_data, kernel = "polynomial")
  svm_models_rbf[[i]] <- svm(Class ~ ., data = train_data, kernel = "radial")
}

```

## Step 4: Evaluating the Models
1. For each of the five train/test splits:
  - Train three SVM models with different kernels:
        -linear
        -polynomial
        -radial
2. predict the corresponding test set.
3. Measure test set accuracy for each model and store the results in a table
```{r evaluate-models}
# Create a results data frame
results <- data.frame(
  Split = integer(),
  Kernel = character(),
  Accuracy = numeric(),
  stringsAsFactors = FALSE
)

# Train and evaluate models for each split
for (i in 1:5) {
  train_data <- train_list[[i]]
  test_data <- test_list[[i]]
  
  # Train models
  svm_linear <- svm(Class ~ ., data = train_data, kernel = "linear")
  svm_poly <- svm(Class ~ ., data = train_data, kernel = "polynomial")
  svm_rbf <- svm(Class ~ ., data = train_data, kernel = "radial")
  
  # Make predictions
  pred_linear <- predict(svm_linear, newdata = test_data)
  pred_poly <- predict(svm_poly, newdata = test_data)
  pred_rbf <- predict(svm_rbf, newdata = test_data)
  
  # Calculate accuracies
  acc_linear <- mean(pred_linear == test_data$Class)
  acc_poly <- mean(pred_poly == test_data$Class)
  acc_rbf <- mean(pred_rbf == test_data$Class)
  
  # Save results
  results <- rbind(results, data.frame(Split = i, Kernel = "Linear", Accuracy = acc_linear))
  results <- rbind(results, data.frame(Split = i, Kernel = "Polynomial", Accuracy = acc_poly))
  results <- rbind(results, data.frame(Split = i, Kernel = "RBF", Accuracy = acc_rbf))
}

# Print the results
print(results)
```

### Observations
Linear kernel is consistently the highest or very close to the highest accuracy across most splits.

RBF sometimes gets close (especially on Split 3 where it barely beats linear), but overall is less consistent.

Polynomial is consistently lower than both Linear and RBF.

Best choice: Linear kernel

## Step 5 : Tune the kernel
1. Tune the linear kernel using the tune() function.
2. Explore cost values: 0.001, 0.01, 0.1, 1, 10, 100.
```{r tune-kernel}
# Choose one training set (e.g., first split)
train_data <- train_list[[1]]

# Tune the cost parameter
set.seed(123)  # Set seed for reproducibility

tune_out <- tune(
  svm,
  Class ~ .,
  data = train_data,
  kernel = "linear",
  ranges = list(cost = c(0.001, 0.01, 0.1, 1, 10, 100))
)

# View the tuning results
summary(tune_out)
```
### Results from tuning:
Best cost: 1

Best performance aka the lowest cross-validation error: ~0.1987 (about 19.87% error)

So... ~80.13% correct classification inside training cross-validation

Cost = 1 was better than very small (0.001, 0.01) or very large (100) cost values.

## Step 6: Final Model Evaluation
Using the tuned parameter (cost = 1), train the final linear SVM model


```{r final-evaluation}
# Create a final results data frame
final_results <- data.frame(
  Split = integer(),
  Accuracy = numeric(),
  stringsAsFactors = FALSE
)

# Final evaluation across all 5 splits
for (i in 1:5) {
  train_data <- train_list[[i]]
  test_data <- test_list[[i]]
  
  # Train SVM with cost = 1
  svm_final <- svm(Class ~ ., data = train_data, kernel = "linear", cost = 1)
  
  # Predict on test set
  predictions <- predict(svm_final, newdata = test_data)
  
  # Accuracy
  acc <- mean(predictions == test_data$Class)
  
  # Save results
  final_results <- rbind(final_results, data.frame(Split = i, Accuracy = acc))
}

# View final results
print(final_results)
```
### Results
The final test set accuracies for each split were:
The models achieved consistently strong performance across 
all splits, with accuracies ranging from 74% to 80%.

The average accuracy across all splits was approximately 78%.

This suggests that the linear SVM with cost = 1 provides a stable 
and effective classification model for the Vehicle dataset.

## Conclusion
I applied Support Vector Machines (SVMs) to the Vehicle dataset to predict vehicle types based
on various measured attributes. I preprocessed data ad created 5 training/test splits. I selected
the linear kernel as the best.

By tuning the linear SVM's cost parameter, I selected cost = 1 as the best value. Using the
tuned model, final test set accuracies ranged from approximately 74% to 80%, with an average final
accuracy around 78%.

These results show that the linear SVM performed effectively on the Vehicle dataset, suggesting
that the underlying feature space is relatively linearly separable. Overall, the linear SVM
provided stable and reliable classification for this problem.

# Higher Education dataset
Load the Higher Education dataset (UCI Machine Learning Repository)
```{r higher-education}
# Load the Higher Education dataset
# Load libraries (again, just in case)
library(e1071)
library(caret)
library(dplyr)

# Load your dataset
# Adjust the file path to where your CSV is saved
edu_data <- read.csv("student-data.csv")

# View the structure of the data
str(edu_data)
summary(edu_data)
sum(is.na(edu_data)) # Check for missing values, we want 0 

# Remove the STUDENT.ID column because it's not useful for prediction
edu_data <- edu_data %>% select(-STUDENT.ID)

```

## Step 1: Data Examination and Preprocessing
```{r edu-preprocess}
# Suppose the target column is called "Output.grade"
# Rename if needed to remove spaces
colnames(edu_data) <- make.names(colnames(edu_data))

# Separate predictors and target
edu_target <- edu_data$GRADE

# Remove target from predictors
edu_predictors <- edu_data %>% select(-GRADE)

# Identify numeric columns
numeric_columns <- sapply(edu_predictors, is.numeric)

# Scale only numeric columns
edu_predictors_scaled <- edu_predictors
edu_predictors_scaled[, numeric_columns] <- scale(edu_predictors[, numeric_columns])

# Combine back predictors + target
edu_processed <- data.frame(edu_predictors_scaled, GRADE = edu_target)
edu_processed$GRADE <- as.factor(edu_processed$GRADE)


# Quick view
head(edu_processed)
```

### Step 1: Data Loading and Preprocessing (Higher Education Dataset)
1. Loaded the Higher Education Students Performance Evaluation dataset from CSV.
2. Renamed columns to valid R variable names using make.names().
3. Selected "GRADE" as the target variable for prediction.
(There were no missing values)
4. Processed predictors

## Step 2: Splitting the Data
```{r edu-split}
set.seed(123)  # Set a seed for reproducibility

# Create empty lists to hold training and testing sets
edu_train_list <- list()
edu_test_list <- list()

# Create 5 different splits
for (i in 1:5) {
  # Make indices for 70% training
  train_indices <- createDataPartition(edu_processed$GRADE, p = 0.7, list = FALSE)
  
  # Split the data
  train_data <- edu_processed[train_indices, ]
  test_data <- edu_processed[-train_indices, ]
  
  # Save them
  edu_train_list[[i]] <- train_data
  edu_test_list[[i]] <- test_data
}

```

### Step 2: Train/Test Set Creation
1. I used createDataPartition() to create 5 different random training/test splits.
2. Each split used 70% of the data for training and 30% for testing.
3. I saved the splits in edu_train_list and edu_test_list for later use.

## Step 3: Training the Models
Plan:
1. Train SVM models with different kernels (linear, polynomial, radial) for each split.
2. Predict on the corresponding test set
3. Record the accuracy for each model
```{r edu-train}
# Create a results data frame to store accuracies
edu_results <- data.frame(
  Split = integer(),
  Kernel = character(),
  Accuracy = numeric(),
  stringsAsFactors = FALSE
)

# Train and evaluate models for each split
for (i in 1:5) {
  train_data <- edu_train_list[[i]]
  test_data <- edu_test_list[[i]]
  
  # 1. Linear kernel
  svm_linear <- svm(GRADE ~ ., data = train_data, kernel = "linear")
  pred_linear <- predict(svm_linear, newdata = test_data)
  acc_linear <- mean(pred_linear == test_data$GRADE)
  edu_results <- rbind(edu_results, data.frame(Split = i, Kernel = "Linear", Accuracy = acc_linear))
  
  # 2. Polynomial kernel
  svm_poly <- svm(GRADE ~ ., data = train_data, kernel = "polynomial")
  pred_poly <- predict(svm_poly, newdata = test_data)
  acc_poly <- mean(pred_poly == test_data$GRADE)
  edu_results <- rbind(edu_results, data.frame(Split = i, Kernel = "Polynomial", Accuracy = acc_poly))
  
  # 3. Radial kernel
  svm_rbf <- svm(GRADE ~ ., data = train_data, kernel = "radial")
  pred_rbf <- predict(svm_rbf, newdata = test_data)
  acc_rbf <- mean(pred_rbf == test_data$GRADE)
  edu_results <- rbind(edu_results, data.frame(Split = i, Kernel = "RBF", Accuracy = acc_rbf))
}

# View the results
print(edu_results)
```

### Step 4: Kernel Selection
1. Kernels: (linear, polynomial, radial) evaluated across five train/test splits.
The polynomial kernel achieved the highest or tied-for-highest accuracy in most splits.
Based on these evaluations, the polynomial kernel is best for parameter tuning and final evaluation.

## Step 5: Tuning the Polynomial Kernel
Details:
1. Chose the first training set for tuning.
2. Tuned the polynomial kernel using a grid search for cost, gamma, and degree.
```{r edu-tune}
# Pick first training set
train_data <- edu_train_list[[1]]

# Tune the polynomial kernel
set.seed(123)  # For reproducibility

tune_out <- tune(
  svm,
  GRADE ~ .,
  data = train_data,
  kernel = "polynomial",
  ranges = list(
    cost = c(0.01, 0.1, 1, 10),
    gamma = c(0.001, 0.01, 0.1, 1),
    degree = c(2, 3, 4)
  )
)

# View tuning results
summary(tune_out)
```

### Step 5: Parameter Tuning
Step 5: Parameter Tuning -> 10-fold cross-validation
Parameters tuned: cost, gamma, and degree.
Explored:
    cost: 0.01, 0.1, 1, 10
    gamma: 0.001, 0.01, 0.1, 1
    degree: 2, 3, 4
Best parameters found:
    cost = 0.1
    gamma = 0.1
    degree = 3

Best cross-validation error: ~75.18%.

## Step 6: Final Model Evaluation
Plan:
For each split
Train a polynomial SVM with the best parameters (cost = 0.1, gamma = 0.1, degree = 3)
Predict on the test set, and record the final accuracy for each split.
```{r edu-final-eval}
# Create a final results data frame
edu_final_results <- data.frame(
  Split = integer(),
  Accuracy = numeric(),
  stringsAsFactors = FALSE
)

# Final evaluation across all 5 splits
for (i in 1:5) {
  train_data <- edu_train_list[[i]]
  test_data <- edu_test_list[[i]]
  
  # Train polynomial SVM with best parameters
  svm_final <- svm(
    GRADE ~ ., 
    data = train_data,
    kernel = "polynomial",
    cost = 0.1,
    gamma = 0.1,
    degree = 3
  )
  
  # Predict on the test set
  predictions <- predict(svm_final, newdata = test_data)
  
  # Calculate accuracy
  acc <- mean(predictions == test_data$GRADE)
  
  # Save results
  edu_final_results <- rbind(edu_final_results, data.frame(Split = i, Accuracy = acc))
}

# View the final results
print(edu_final_results)
```

### Step 6: Evaluation and Analysis
The final models were trained on each of the 5 training/test splits using the tuned polynomial SVM (parameters: cost = 0.1, gamma = 0.1, degree = 3).

Final test set accuracies for each split were:
Split	Accuracy
1	26.83%
2	24.39%
3	24.39%
4	26.83%
5	29.27%

The average final test set accuracy across all splits was approximately 26%.

## Conclusion
I explored the use of Support Vector Machines (SVMs) with different kernels 
(linear, polynomial, and radial) to classify student performance based on various 
academic and demographic features. After evaluating models across five training/test 
splits, the polynomial kernel was selected as the best-performing option.

Using parameter tuning, the best polynomial SVM was found with parameters: 
cost = 0.1, gamma = 0.1, and degree = 3. 
Final evaluation across all five data splits resulted in test accuracies ranging 
from approximately 24% to 29%, with an overall average accuracy of about 26%.

While the tuned polynomial SVM showed slightly better performance than random guessing, 
the overall accuracy suggests that the dataset presents challenges for classification. 
Possible factors impacting performance include:
- Limited predictive power in the available features
- Potential imbalance or noise in the dataset

For future work, improvements could involve:
- Engineering more informative features
- Trying alternative models like decision trees and random forests
- Balancing the classes if severe class imbalance exists

